---
title: "P8105 HW3"
author: "Yijin Serena Wang"
date: "`r Sys.Date()`"
output: github_document
---

```{r, message = FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(ggplot2)
library(p8105.datasets)
library(patchwork)
data("ny_noaa")
data("instacart")
```

```{r graph settings}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1
The `instacart` data set contains users' online purchasing information in 2017. It has `r nrow(instacart)` rows and `r ncol(instacart)` columns. The dataset is organized by `order_id` and `product_id` of each user. It also has detailed information for each order such as `order_dow`, `days_since_prior_order`. 

illustrative examples

```{r aisles}
aisle_product_count <- instacart %>% 
  select(aisle, product_id) %>%
  group_by(aisle) %>%
  summarize(product_count = n()) %>%
  arrange(desc(product_count))

head(aisle_product_count)

```
There are `r nrow(aisle_product_count)` aisles. Fresh vegetables has the most items ordered. 

```{r aisle product plot}
aisle_product_count %>%
  filter(product_count > 10000) %>%
  ggplot(aes(x = reorder(aisle, -product_count), y = product_count)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) + 
  labs(x = "Aisle", y = "Number of items ordered")

```
```{r}
instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  select(aisle, product_name) %>%
  group_by(aisle,product_name) %>%
  summarise(product_count = n()) %>%
  top_n(3,product_count) %>%
  ungroup() %>%
  knitr::kable(digits = 1)
  
```

```{r table}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream"))  %>%
  select(order_hour_of_day, order_dow, product_name) %>%
  group_by(product_name, order_dow) %>%
  summarise(mean_order_hour = mean(order_hour_of_day)) %>%
  ungroup()  %>%
  pivot_wider(names_from = product_name,
              values_from = mean_order_hour) %>%
  knitr::kable(digits = 1)

```

# Problem 2
```{r}
# load and clean accelerometers data set
accel_data <- read_csv("./Data/accel_data.csv") %>%
  # pivot activity.* columns longer
  # minute column has the minute number associated with activity
  # activity column has corresponding values to the original activity.* columns
  pivot_longer(
    cols = starts_with("activity"),
    names_to = "minute",
    names_prefix = "activity.",
    values_to ="activity"
  ) %>%
  # clean names
  clean_names() %>%
  # create a column to denote if the day value is weekday or weekend
  mutate(is_weekday = ifelse(
    day %in% c("Saturday", "Sunday"), 
    FALSE, 
    TRUE))
```

The resulting data set has `r nrow(accel_data)` rows and `r ncol(accel_data)` columns. It contains `r colnames(accel_data)` columns. It contains 5 weeks of data of an accelerometer worn by a patient.

```{r}
accel_data %>% 
  group_by(day_id) %>%
  summarise(total_activity = sum(activity)) %>%
  knitr::kable(digits = 1)
  

```
The trends are not apparent. 

```{r}
ggplot(data = accel_data, 
       aes(x = minute, y = activity, color = day)) +
  geom_point(alpha = .5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "24-hour activity time courses")
```

It seems for every day in a week, there is a specific period of time that has low activity. At other times of the day, there is a day-effect on the time that the participant did activity. There are dots that are on top of one another, forming a straight line. These dots are all in the same color, indicating the same day of the week. In general, the participant did more activity in the morning (assumingly) on Fridays and Saturdays, in later of the day (assumingly) on Thursdays and Tuesdays.


# Problem 3

The `ny_noaa` data set has `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. It contains precipitation, snowfall, depth of snow and min and max temperatures of each documented NY state weather stations from `r min(ny_noaa$date)` and `r max(ny_noaa$date)`. Just by looking at the data, there are a lot of cells with NAs. Sometimes, there is no information about precipitation, snowfall, depth of snow and min and max temperatures for a particular station and a date. min and max temperatures are missing for a lot of stations and dates.
```{r cleaning}
clean_ny_noaa <- ny_noaa %>%
  # clean names
  clean_names() %>%
  # break date column into month, day, year columns
  separate(col = date, 
           sep = "-",
           into = c("year", "month", "day")) %>%
  # convert precipitation to mm, temperatures to degree Celcius 
  # and snow depth to cm
  mutate(prcp = prcp/10,
         snwd = snwd/10,
         tmax = as.numeric(tmax)/10,
         tmin = as.numeric(tmin)/10)
```

```{r}
clean_ny_noaa %>%
  count(snow, name = "count") %>%
  top_n(1, count) %>%
  knitr::kable(digits = 1)
  
```
The most commonly observed value is 0. It might be because the actual data is documented in the `snwd` column. 

```{r avg max temp plot}
clean_ny_noaa  %>%
  filter(month %in% c("01","07"))  %>%
  group_by(year, month, id) %>%
  # calculate average of max temperatures for observations in each station in each month, year
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>%
  ungroup() %>%
  # make plot
  ggplot(aes(x = year, y = avg_tmax, group = id)) +
  geom_line(alpha = .2) +
  # create facets based on month
  facet_wrap(. ~ month) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  labs(title = "Average max temperature for each station", y = "average max temperature" )
```
It seems like the average temperatures in July are all higher than those in January across these years. Average temperatures in July seem to have a lower variance than those in January. Also, it seems that for both January and July, the average max temperatures has similar fluctuation around the same years. There is one significant outlier in approximately 1988 July.  

```{r temp and snowfall plot}
tmax_tmin_p <- clean_ny_noaa %>%
  pivot_longer(cols = c("tmax", "tmin"), 
               names_to = "temp_type", 
               values_to = "temp_value")  %>%
  ggplot(aes(x = temp_type, y = temp_value)) +
  geom_boxplot() + 
  labs(title = "Max Temperature vs. Min Temperature")
  

snowfall_p <- clean_ny_noaa %>%
  filter(snow < 100 & snow > 0) %>%
  ggplot(aes(x = snow, fill = year)) + 
  geom_density(alpha = .3) +
  theme(legend.position = "right") +
  labs(title = "Distribution of snowfall", x = "snowfall")

tmax_tmin_p/snowfall_p
  
```





