---
title: "P8105 HW3"
author: "Yijin Serena Wang"
date: "`r Sys.Date()`"
output: github_document
---

```{r, message = FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(ggplot2)
library(p8105.datasets)
library(patchwork)
library(ggridges)
data("ny_noaa")
data("instacart")
```

```{r graph settings}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1
The `instacart` data set contains users' online purchasing information in 2017. It has `r nrow(instacart)` rows and `r ncol(instacart)` columns. The dataset is organized by `order_id` and `product_id` of each user. It also has detailed information for each order such as `order_dow`, `days_since_prior_order`. 

```{r aisles}
aisle_product_count <- instacart %>%
  select(aisle, product_id) %>%
  group_by(aisle) %>%
  # calculate count
  summarize(product_count = n()) %>%
  arrange(desc(product_count))

```
There are `r nrow(aisle_product_count)` aisles. Fresh vegetables has the most items ordered. 

```{r aisle product plot}
aisle_product_count %>%
  # filter data
  filter(product_count > 10000) %>%
  # create plot
  ggplot(aes(x = reorder(aisle,-product_count), y = product_count)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  # add labels
  labs(x = "Aisle", y = "Number of items ordered")

```
```{r}
instacart %>%
  # filter data
  filter(aisle %in% c(
    "baking ingredients",
    "dog food care",
    "packaged vegetables fruits"
  )) %>%
  select(aisle, product_name) %>%
  group_by(aisle, product_name) %>%
  # calculate count
  summarise(product_count = n()) %>%
  # get top 3
  top_n(3, product_count) %>%
  ungroup() %>%
  knitr::kable(digits = 1)

```

```{r table}
instacart %>%
  # filter data
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream"))  %>%
  select(order_hour_of_day, order_dow, product_name) %>%
  group_by(product_name, order_dow) %>%
  # calculate mean
  summarise(mean_order_hour = mean(order_hour_of_day)) %>%
  ungroup()  %>%
  # pivot to a table
  pivot_wider(names_from = product_name,
              values_from = mean_order_hour) %>%
  knitr::kable(digits = 1)

```

# Problem 2
```{r}
# load and clean accelerometers data set
accel_data <- read_csv("./Data/accel_data.csv") %>%
  # clean names
  clean_names() %>%
  # pivot activity_* columns longer
  # minute column has the minute number associated with activity
  # activity column has corresponding values to the original activity.* columns
  pivot_longer(
    cols = starts_with("activity_"),
    names_to = "minute",
    names_prefix = "activity_",
    values_to = "activity"
  ) %>%
  mutate(
    # create a column to denote if the day value is weekday or weekend
    day_type = ifelse(day %in% c("Saturday", "Sunday"),
                             "weekend",
                             "weekday"),
    # make day factors for calculating total activities later
    day = factor(
      day,
      levels = c(
        "Monday",
        "Tuesday",
        "Wednesday",
        "Thursday",
        "Friday",
        "Saturday",
        "Sunday"
      )
    ),
    # convert minute to numbers
    minute = as.integer(minute))
```

The resulting data set has `r nrow(accel_data)` rows and `r ncol(accel_data)` columns. It contains `r colnames(accel_data)` columns. It contains 5 weeks of data of an accelerometer worn by a patient. The data is documented at each minute interval in 24 hours for each day.

```{r}
accel_data %>%
  group_by(day) %>%
  # calculate total activity
  summarise(total_activity = sum(activity)) %>%
  knitr::kable(digits = 1)


```
There are more activities from Wednesday to Friday. Other days of the week have similar amount of activities.  

```{r}
# create plot
ggplot(data = accel_data,
  aes(x = minute, y = activity, color = day)) +
  geom_line(alpha = .3) +
  # rotate x-axis labels
  theme(axis.text.x = element_text(
    angle = 90,
    vjust = 0.5,
    hjust = 1
  )) +
  # add title
  labs(title = "24-hour activity time courses") + 
  # move legend to the right
  theme(legend.position = "right")
  
```

Assume min 1 is midnight. It seems for every day in a week, there is a specific period of time (might be mornings) that has low activity. At other times of the day, there is a day-effect on the time that the participant did activity. There a cluster of yellow line for Sundays at centered at a period of time and some clusters for days in the early part of the week. In general, the participant did more activity in  very-early morning and evenings on of weekdays, more activity in the late morning and afternoon at weekends.


# Problem 3

```{r cleaning}
clean_ny_noaa <- ny_noaa %>%
  # clean names
  clean_names() %>%
  # break date column into month, day, year columns
  separate(col = date,
           sep = "-",
           into = c("year", "month", "day")) %>%
  # convert precipitation to mm, temperatures to degree Celcius
  # and snow depth to cm
  mutate(
    prcp = prcp / 10,
    snwd = snwd / 10,
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10
  )
```

The `ny_noaa` data set has `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. It contains precipitation, snowfall, depth of snow and min and max temperatures of each documented NY state weather stations from `r min(ny_noaa$date)` and `r max(ny_noaa$date)`. Just by looking at the data, there are a lot of cells with NAs. Sometimes, there is no information about precipitation, snowfall, depth of snow and min and max temperatures for a particular station and a date. min and max temperatures are missing for a lot of stations and dates.

The cleaned data set has `r nrow(clean_ny_noaa)` rows and `r ncol(clean_ny_noaa)` columns. 

```{r}
clean_ny_noaa %>%
  # count values
  count(snow, name = "count") %>%
  # take the top one
  top_n(1, count) %>%
  knitr::kable(digits = 1)
  
```
The most commonly observed value is 0. It might be because the actual data is documented in the `snwd` column. Or it could also be another way for documenting missing values, instead of using NAs.

```{r avg max temp plot}
clean_ny_noaa  %>%
  # filter data
  filter(month %in% c("01", "07"))  %>%
  group_by(year, month, id) %>%
  # calculate average of max temperatures for observations in each station in each month, year
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>%
  ungroup() %>%
  # make plot
  ggplot(aes(x = year, y = avg_tmax, group = id)) +
  geom_line(alpha = .2) +
  # create facets based on month
  facet_wrap(. ~ month) +
  theme(axis.text.x = element_text(
    angle = 90,
    vjust = 0.5,
    hjust = 1
  ),
  # remote legends
  legend.position = "none") +
  # add title and label
  labs(title = "Average max temperature for each station", y = "average max temperature")
```

It seems like the average temperatures in July are all higher than those in January across these years. Average temperatures in July seem to have a lower variance than those in January. Also, it seems that for both January and July, the average max temperatures has similar fluctuation around the same years. There is one significant outlier in approximately 1988 July.  

```{r temp and snowfall plot, out.width="100%"}
tmax_tmin_p <- clean_ny_noaa %>%
  # reformat data
  pivot_longer(
    cols = c("tmax", "tmin"),
    names_to = "temp_type",
    values_to = "temp_value"
  )  %>%
  # create boxplot
  ggplot(aes(x = temp_type, y = temp_value)) +
  geom_boxplot() +
  labs(title = "Max Temperature vs. Min Temperature",
       plot.title = element_text(size=8))


snowfall_p <- clean_ny_noaa %>%
  # filter data
  filter(snow < 100 & snow > 0) %>%
  # create ridge plot
  ggplot(aes(x = snow, y = year)) +
  geom_density_ridges(alpha = .3) +
  # move legend to the right
  theme(legend.position = "right") +
  # add title and label
  labs(title = "Distribution of snowfall",
       x = "snowfall",
       plot.title = element_text(size = 8))

tmax_tmin_p + snowfall_p
  
```




